---
title: "Data Exercise"
---

First, I will create a synthetic dataset with multiple variables and some associations between them. I will then explore the data with plots and tables, and fit some simple models to see if they can recover the associations.

### Generate Synthetic Dataset

```{r}
# Load necessary libraries
library(tidyr)
library(dplyr)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Generate synthetic dataset
n <- 1000
x1 <- rnorm(n, mean = 5, sd = 2)
x2 <- rnorm(n, mean = 3, sd = 1)
x3 <- rbinom(n, 1, 0.5)
y <- 3 + 2 * x1 - 1.5 * x2 + 0.8 * x3 + rnorm(n)

# Create data frame
df <- data.frame(y, x1, x2, x3)

# Explore the dataset
summary(df)
head(df)
```

### Explore the Data

#### Plot the relationships between variables

```{r}
# Plot y against x1
ggplot(df, aes(x = x1, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Scatter plot of y vs x1")

# Plot y against x2
ggplot(df, aes(x = x2, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Scatter plot of y vs x2")

# Boxplot of y by x3
ggplot(df, aes(x = as.factor(x3), y = y)) +
  geom_boxplot() +
  labs(title = "Boxplot of y by x3", x = "x3")
```

### Fit Simple Models

#### Linear Model

```{r}
# Fit a linear model
model <- lm(y ~ x1 + x2 + x3, data = df)
summary(model)
```

#### Explore Model Coefficients

```{r}
# Print the coefficients of the linear model
cat("Coefficients of the linear model:\n")
print(coef(model))
```

### Explore Different Models

#### Polynomial Regression

```{r}
# Fit a polynomial regression model
model_poly <- lm(y ~ poly(x1, 2) + poly(x2, 2) + x3, data = df)
summary(model_poly)
```

#### Interaction Model

```{r}
# Fit a model with interaction terms
model_interact <- lm(y ~ x1 * x2 + x1 * x3 + x2 * x3, data = df)
summary(model_interact)
```

### Conclusion

This R code generates a synthetic dataset with multiple variables and known associations, explores the data with plots and tables, and fits several simple models to recover the associations. The linear model should reveal the coefficients that were built into the data generation process, while the polynomial and interaction models allow us to explore more complex relationships.
