---
title: "CDC Data Exploration on Weekly United States Hospitalization Metrics by Jurisdiction"
author: "Seth Harris"
date: "2024-07-05"
format: html
editor: visual
---

# Brief Description of the Data

**Title:** Weekly United States Hospitalization Metrics by Jurisdiction, During Mandatory Reporting Period from August 1, 2020 to April 30, 2024, and for Data Reported Voluntarily Beginning May 1, 2024, National Healthcare Safety Network (NHSN)

**Source:** (https://data.cdc.gov/Public-Health-Surveillance/Weekly-United-States-Hospitalization-Metrics-by-Ju/aemt-mg7g/about_data)

**Description:** This dataset provides information on COVID-19 and influenza-related hospitalizations, hospital occupancy, and hospital capacity across different jurisdictions in the United States from August 1, 2020, to April 30, 2024, and data reported voluntarily from May 1, 2024. It includes both continuous and categorical variables. I will further explore the dataset from the CDC data website. Specifically looking at information on hospitalization metrics, including the number of new admissions, total number of beds occupied, and more.

# Loading and Processing the Data

**The first step is to load the data and perform some initial cleaning and processing.**

```{r}
# Load necessary libraries 
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r}
# Load the dataset
cdcdata <- read_csv("https://data.cdc.gov/api/views/aemt-mg7g/rows.csv?accessType=DOWNLOAD")
```

```{r}
# Inspect the first few rows of the data and column names
head(cdcdata)
colnames(cdcdata)
```

```{r}
# Check for missing values
summary(cdcdata)
```

**Next, I will process the data: Handle any missing values or non-standard codes and select relevant columns for analysis.**

```{r}
# Convert date column to Date type
cdcdata <- cdcdata %>%
  mutate(`Week Ending Date` = as.Date(`Week Ending Date`, format = "%Y-%m-%d")) %>%
  mutate_all(~ifelse(. == 999, NA, .))

# Select relevant columns for analysis
selected_data <- cdcdata %>%
  select(`Week Ending Date`, `Geographic aggregation`, `Number Hospitals Reporting Adult COVID-19 Admissions`, `Weekly Average Percent COVID-19 Inpatient Bed Occupancy`, `Weekly Average COVID-19 Admissions`, `Percent Hospitals Reporting Inpatient Beds Occupied`, `Weekly Average Inpatient Beds Occupied`, `Weekly Average ICU Beds Occupied`)

# Inspect the cleaned data
head(selected_data)
summary(selected_data)
```

# Exploratory Data Analysis

**In this section, I will perform some exploratory/descriptive analysis on this cleaned dataset.**

```{r}
# Summary table of key variables
summary(selected_data)
```

```{r}
# Plot the distribution of Weekly Average COVID-19 Admissions
ggplot(selected_data, aes(x = `Weekly Average COVID-19 Admissions`)) +
  geom_histogram(binwidth = 500, fill = "blue", color = "blue") +
  labs(title = "Distribution of Weekly Average COVID-19 Admissions", x = "Weekly Average COVID-19 Admissions", y = "Frequency")
```

This histogram shows the distribution of Weekly Average COVID-19 Admissions across the different jurisdictions. The majority of jurisdictions report fewer than 2000 admissions per week, with a long tail indicating that a few jurisdictions have significantly higher weekly admissions.

I can see a very high frequency of lower values with a significant right skew. Most of the data points are clustered around the lower range of weekly average COVID-19 admissions, with a few outliers having very high values. This indicates that most jurisdictions have relatively low weekly average COVID-19 admissions, while a few have extremely high numbers.

```{r}
# Plot the distribution of Weekly Average Inpatient Beds Occupied
ggplot(selected_data, aes(x = `Weekly Average Inpatient Beds Occupied`)) +
  geom_histogram(binwidth = 500, fill = "black", color = "red") +
  labs(title = "Distribution of Weekly Average Inpatient Beds Occupied", x = "Weekly Average Inpatient Beds Occupied", y = "Frequency")
```

This histogram shows the distribution of Weekly Average Inpatient Beds Occupied. Similar to the previous plot, most jurisdictions have fewer than 2000 occupied inpatient beds per week, with a few jurisdictions having significantly higher numbers.

Similar to the previous histogram, this one also exhibits a strong right skew. Most jurisdictions have a lower number of weekly average inpatient beds occupied, with a few outliers at the higher end. The concentration of data points at the lower end suggests that most regions did not see a large number of inpatient beds occupied weekly.

```{r}
# Plot the distribution of Weekly Average ICU Beds Occupied
ggplot(selected_data, aes(x = `Weekly Average ICU Beds Occupied`)) +
  geom_histogram(binwidth = 500, fill = "lightblue", color = "lightblue") +
  labs(title = "Distribution of Weekly Average ICU Beds Occupied", x = "Weekly Average ICU Beds Occupied", y = "Frequency")
```

This histogram shows the distribution of Weekly Average ICU Beds Occupied. The distribution is skewed towards lower numbers, with most jurisdictions reporting fewer than 500 ICU beds occupied per week.

This histogram also shows a right-skewed distribution, indicating that the majority of jurisdictions have a lower weekly average of ICU beds occupied. The distribution is similar to that of inpatient beds but with ICU-specific data.

```{r}
# Summary of categorical variable 'Geographic aggregation'
geo_summary <- selected_data %>%
  group_by(`Geographic aggregation`) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

geo_summary
```

This summary table shows the count of data points for each Geographic aggregation. This helps us understand the distribution of the data across different jurisdictions.

```{r}
# Summary of categorical variable 'Geographic aggregation'
geo_summary <- selected_data %>%
  group_by(`Geographic aggregation`) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Filter for top 5 and bottom 5 states by count
top_5_states <- geo_summary %>% top_n(5, count)
bottom_5_states <- geo_summary %>% top_n(-5, count)

# Combine top and bottom 5 states
selected_states <- bind_rows(top_5_states, bottom_5_states)

# Filter the data for these states
filtered_data <- selected_data %>% 
  filter(`Geographic aggregation` %in% selected_states$`Geographic aggregation`)

# Plot the percentage of data points per Geographic aggregation for selected states
ggplot(filtered_data, aes(x = reorder(`Geographic aggregation`, -`Weekly Average COVID-19 Admissions`), y = `Weekly Average COVID-19 Admissions`)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  labs(title = "Weekly Average COVID-19 Admissions for Selected States", x = "Geographic Aggregation", y = "Weekly Average COVID-19 Admissions")
```

This bar plot shows the Weekly Average COVID-19 Admissions for the top 5 and bottom 5 states. This helps us identify states with the highest and lowest average admissions.

This bar plot depicts the weekly average COVID-19 admissions for selected states. The plot highlights the stark contrast between states with very high and very low average COVID-19 admissions. It visually demonstrates that while most states have low to moderate admissions, a few have extremely high numbers.

```{r}
# Create boxplot for Weekly Average Percent COVID-19 Inpatient Bed Occupancy
ggplot(filtered_data, aes(x = `Geographic aggregation`, y = `Weekly Average Percent COVID-19 Inpatient Bed Occupancy`)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  coord_flip() +
  labs(title = "Boxplot of Weekly Average Percent COVID-19 Inpatient Bed Occupancy by State", x = "Geographic Aggregation", y = "Weekly Average Percent COVID-19 Inpatient Bed Occupancy")
```

This boxplot shows the distribution of Weekly Average Percent COVID-19 Inpatient Bed Occupancy across the selected states. It provides insights into the central tendency and variability of bed occupancy in these states.

This boxplot reveals the variation in the weekly average percent COVID-19 inpatient bed occupancy across different states. Some states have a higher median occupancy rate and more spread out data points, indicating greater variability in bed occupancy percentages.

```{r}
# Create boxplot for Percent Hospitals Reporting Inpatient Beds Occupied
ggplot(filtered_data, aes(x = `Geographic aggregation`, y = `Percent Hospitals Reporting Inpatient Beds Occupied`)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  coord_flip() +
  labs(title = "Boxplot of Percent Hospitals Reporting Inpatient Beds Occupied by State", x = "Geographic Aggregation", y = "Percent Hospitals Reporting Inpatient Beds Occupied")
```

This boxplot shows the distribution of Percent Hospitals Reporting Inpatient Beds Occupied across the selected states. It helps us understand the proportion of hospitals reporting occupied beds in each state.

This boxplot shows the percent of hospitals reporting inpatient beds occupied across different states. Most states have a high percentage of hospitals reporting inpatient beds occupied, close to 100%, with some variability. This suggests that hospital bed occupancy reporting is consistent across most jurisdictions.

```{r}
# Create boxplot for Weekly Average Inpatient Beds Occupied
ggplot(filtered_data, aes(x = `Geographic aggregation`, y = `Weekly Average Inpatient Beds Occupied`)) +
  geom_boxplot(fill = "lightcoral", color = "darkred") +
  coord_flip() +
  labs(title = "Boxplot of Weekly Average Inpatient Beds Occupied by State", x = "Geographic Aggregation", y = "Weekly Average Inpatient Beds Occupied")
```

This boxplot shows the distribution of Weekly Average Inpatient Beds Occupied across the selected states. It provides insights into the central tendency and variability of inpatient bed occupancy in these states.

This boxplot indicates the weekly average number of inpatient beds occupied by state. There is a noticeable variation between states, with some having a significantly higher number of occupied beds on average. The presence of outliers suggests that certain states experienced unusually high occupancy levels.

```{r}
# Create boxplot for Weekly Average ICU Beds Occupied
ggplot(filtered_data, aes(x = `Geographic aggregation`, y = `Weekly Average ICU Beds Occupied`)) +
  geom_boxplot(fill = "lightpink", color = "darkred") +
  coord_flip() +
  labs(title = "Boxplot of Weekly Average ICU Beds Occupied by State", x = "Geographic Aggregation", y = "Weekly Average ICU Beds Occupied")
```

This boxplot shows the distribution of Weekly Average ICU Beds Occupied across the selected states. It provides insights into the central tendency and variability of ICU bed occupancy in these states.

Similar to the inpatient beds occupied boxplot, this one shows the weekly average number of ICU beds occupied by state. The variation among states is evident, with some states having higher averages and others lower. The outliers indicate states with exceptionally high ICU bed occupancy.

# Conclusion:

Overall, these plots illustrate the significant variability in COVID-19-related hospital metrics across different jurisdictions. The data is heavily skewed in many cases, indicating that while most areas experienced moderate levels, a few regions faced extremely high demands on their healthcare systems. The boxplots highlight the differences in hospital bed occupancy and reporting consistency across states, providing valuable insights into the impact of COVID-19 on healthcare resources.

This section contributed by **Sri Lakshmi Sudha Ganni**

# Load necessary libraries:

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(faux)
```

# Set the seed for reproducibility:

```{r}
set.seed(123)
```

# Generate synthetic data:

```{r}
n <- 500  # Number of data points
geo_aggregations <- c("State1", "State2", "State3", "State4", "State5", "State6", "State7", "State8", "State9", "State10")

synthetic_data <- data.frame(
  `Week Ending Date` = sample(seq(as.Date('2020-08-01'), as.Date('2024-04-30'), by="week"), n, replace = TRUE),
  `Geographic aggregation` = sample(geo_aggregations, n, replace = TRUE),
  `Number Hospitals Reporting Adult COVID-19 Admissions` = round(runif(n, min = 5, max = 100)),
  `Weekly Average Percent COVID-19 Inpatient Bed Occupancy` = round(runif(n, min = 0, max = 100), 2),
  `Weekly Average COVID-19 Admissions` = round(rnorm(n, mean = 1000, sd = 500)),
  `Percent Hospitals Reporting Inpatient Beds Occupied` = round(runif(n, min = 0, max = 100), 2),
  `Weekly Average Inpatient Beds Occupied` = round(rexp(n, rate = 0.001)),  # Right-skewed distribution
  `Weekly Average ICU Beds Occupied` = round(rnorm(n, mean = 500, sd = 200))
)

```

# Inspect the synthetic data:

```{r}
head(synthetic_data)
summary(synthetic_data)
```
Perform Exploratory Data Analysis (EDA):

# Summary table of key variables:

```{r}
summary(synthetic_data)
```

# Plot the distribution of Weekly Average COVID-19 Admissions:

```{r}
ggplot(synthetic_data, aes(x = `Weekly.Average.COVID.19.Admissions`)) +
  geom_histogram(binwidth = 100, fill = "blue", color = "blue") +
  labs(title = "Distribution of Weekly Average COVID-19 Admissions", x = "Weekly Average COVID-19 Admissions", y = "Frequency")+xlim(0, 50000)
```

Description of the Plot:

The histogram shown in the plot visualizes the distribution of Weekly Average COVID-19 Admissions from the synthetic dataset. Here are the key observations:

X-axis: The Weekly Average COVID-19 Admissions, ranging from 0 to 50,000.

Y-axis: The Frequency, indicating the number of jurisdictions reporting particular ranges of weekly average COVID-19 admissions.

The distribution is highly right-skewed, with most jurisdictions reporting relatively low weekly average COVID-19 admissions.

There is a significant concentration of data points on the lower end of the spectrum, with the majority of the frequencies clustered around the lower values.

The long tail to the right indicates the presence of outliers, jurisdictions with exceptionally high weekly average COVID-19 admissions. 
Interpretation.

The histogram effectively illustrates the disparity in COVID-19 impact across different jurisdictions, showing that while most areas experienced lower average admissions, a few regions faced extremely high admissions, pointing to potential hotspots or areas with larger outbreaks during the reporting period. This pattern can inform public health officials and policymakers about resource allocation and preparedness strategies.

# Plot the distribution of Weekly Average Inpatient Beds Occupied:

```{r}
ggplot(synthetic_data, aes(x = `Weekly.Average.Percent.COVID.19.Inpatient.Bed.Occupancy`)) +
  geom_histogram(fill = "blue", color = "red") +
  labs(title = "Distribution of Weekly Average Inpatient Beds Occupied", x = "Weekly Average Inpatient Beds Occupied", y = "Frequency")+xlim(0, 200)
```
Description of the Plot:

The histogram shown in the plot visualizes the distribution of Weekly Average Inpatient Beds Occupied from the synthetic dataset. Here are the key observations:

1.X-axis: The Weekly Average Inpatient Beds Occupied, ranging from 0 to 200.

2.Y-axis: The Frequency, indicating the number of jurisdictions reporting particular ranges of weekly average inpatient beds occupied.

This histogram of the synthetic data for Weekly Average Inpatient Beds Occupied provides insights into how inpatient bed occupancy is distributed across different jurisdictions. The relatively even distribution and multiple peaks could indicate varying levels of hospital bed occupancy across regions, possibly influenced by different factors such as the population size, healthcare capacity, and severity of outbreaks.The histogram should show a high concentration of lower values with a long tail extending to the right, indicating a few very high values.

This distribution highlights that while some jurisdictions have lower inpatient bed occupancy, others experience higher levels, showcasing the diversity in hospital resource utilization.

# Plot the distribution of Weekly Average ICU Beds Occupied:

```{r}
ggplot(synthetic_data, aes(x = `Weekly.Average.ICU.Beds.Occupied`)) +
  geom_histogram(binwidth = 50, fill = "lightblue", color = "lightblue") +
  labs(title = "Distribution of Weekly Average ICU Beds Occupied", x = "Weekly Average ICU Beds Occupied", y = "Frequency")+xlim(0, 20000)
```
Description of the Plot:

The histogram shown in the plot visualizes the distribution of Weekly Average ICU Beds Occupied from the synthetic dataset.

1.X-axis: The Weekly Average ICU Beds Occupied.

2.Y-axis: The Frequency, indicating the number of jurisdictions reporting particular ranges of weekly average ICU beds occupied.

This histogram exhibits a clear right skew. The majority of the data points are concentrated at the lower end of the range (close to 0), with a long tail extending to the right.

This indicates that the majority of jurisdictions have low to moderate weekly average ICU bed occupancy, but a few jurisdictions experience much higher occupancy levels. This pattern suggests that while most areas manage with fewer occupied ICU beds, certain regions face significant demand, potentially due to higher population densities or more severe outbreaks.



# Summary of categorical variable 'Geographic aggregation':

```{r}
geo_summary <- synthetic_data %>%
  group_by(`Geographic.aggregation`) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

geo_summary
```

# Filter for top 5 and bottom 5 states by count:

```{r}
top_5_states <- geo_summary %>% top_n(5, count)
bottom_5_states <- geo_summary %>% top_n(-5, count)
```

# Combine top and bottom 5 states:

```{r}
selected_states <- bind_rows(top_5_states, bottom_5_states)
```

# Filter the data for these states:

```{r}
filtered_data <- synthetic_data %>% 
  filter(`Geographic.aggregation` %in% selected_states$`Geographic.aggregation`)
```

# Plot the percentage of data points per Geographic aggregation for selected states:

```{r}
ggplot(filtered_data, aes(x = reorder(`Geographic.aggregation`, -`Weekly.Average.COVID.19.Admissions`), y = `Weekly.Average.COVID.19.Admissions`)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  labs(title = "Weekly Average COVID-19 Admissions for Selected States", x = "Geographic Aggregation", y = "Weekly Average COVID-19 Admissions")+ylim(0, 100000)
```
Description of the Plot:

The bar plot visualizes the Weekly Average COVID-19 Admissions for selected states. Here are the key observations:

1. X-axis: The Weekly Average COVID-19 Admissions.

2.Y-axis: The Geographic Aggregation, representing different selected states.

This bar plot provides a clear visual comparison of the weekly average COVID-19 admissions across different states. It highlights the states with the highest and lowest average admissions, which can be useful for identifying areas with greater healthcare needs and resource allocation.

State10: The high average admissions suggest that this state experienced a particularly severe impact from COVID-19, necessitating more healthcare resources and support.

State1: The lower admissions indicate relatively less strain on the healthcare system in this state compared to others.


# Create boxplot for Weekly Average Percent COVID-19 Inpatient Bed Occupancy:

```{r}
ggplot(filtered_data, aes(x = `Geographic.aggregation`, y = `Weekly.Average.Percent.COVID.19.Inpatient.Bed.Occupancy`)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  coord_flip() +
  labs(title = "Boxplot of Weekly Average Percent COVID-19 Inpatient Bed Occupancy by State", x = "Geographic Aggregation", y = "Weekly Average Percent COVID-19 Inpatient Bed Occupancy")+ylim(0, 100)
```
Description of the Plot:

The boxplot visualizes the Weekly Average Percent COVID-19 Inpatient Bed Occupancy for selected states. Here are the key observations:

1.X-axis: The Weekly Average Percent COVID-19 Inpatient Bed Occupancy.

2.Y-axis: The Geographic Aggregation, representing different selected states.

This boxplot provides a clear visual comparison of the weekly average percent COVID-19 inpatient bed occupancy across different states. It highlights the differences in central tendency (median), spread (IQR), and overall range (whiskers) of bed occupancy percentages, indicating how various states managed their inpatient bed capacities during the reporting period.

States with High Occupancy Rates: States like State9 have higher median occupancy rates, suggesting higher utilization of inpatient beds for COVID-19 patients.
States with Lower Occupancy Rates: States like State1 have lower median occupancy rates, indicating relatively less strain on their inpatient bed capacity.



# Create boxplot for Percent Hospitals Reporting Inpatient Beds Occupied:

```{r}
ggplot(filtered_data, aes(x = `Geographic.aggregation`, y = `Percent.Hospitals.Reporting.Inpatient.Beds.Occupied`)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  coord_flip() +
  labs(title = "Boxplot of Percent Hospitals Reporting Inpatient Beds Occupied by State", x = "Geographic Aggregation", y = "Percent Hospitals Reporting Inpatient Beds Occupied")
```
Description of the Plot:

The boxplot visualizes the Percent Hospitals Reporting Inpatient Beds Occupied by state. Here are the key observations:

1.X-axis: The Percent Hospitals Reporting Inpatient Beds Occupied.

2.Y-axis: The Geographic Aggregation, representing different selected states.

This boxplot provides a clear visual comparison of the percentage of hospitals reporting inpatient beds occupied across different states. It highlights the differences in central tendency (median), spread (IQR), and overall range (whiskers) of reporting percentages, indicating how various states managed their inpatient bed reporting during the reporting period.

States with High Reporting Percentages: States like State10 have higher median reporting percentages, suggesting higher compliance or higher utilization of inpatient beds.

States with Lower Reporting Percentages: States like State1 have lower median reporting percentages, indicating relatively less strain on their inpatient bed capacity or lower compliance in reporting.


# Create boxplot for Weekly Average Inpatient Beds Occupied:

```{r}
ggplot(filtered_data, aes(x = `Geographic.aggregation`, y = `Weekly.Average.Inpatient.Beds.Occupied`)) +
  geom_boxplot(fill = "lightcoral", color = "darkred") +
  coord_flip() +
  labs(title = "Boxplot of Weekly Average Inpatient Beds Occupied by State", x = "Geographic Aggregation", y = "Weekly Average Inpatient Beds Occupied")
```
Description of the Plot:

The boxplot visualizes the Weekly Average Inpatient Beds Occupied by state. Here are the key observations:

1.X-axis: The Weekly Average Inpatient Beds Occupied.

2.Y-axis: The Geographic Aggregation, representing different selected states.

This boxplot provides a clear visual comparison of the weekly average inpatient beds occupied across different states. It highlights the differences in central tendency (median), spread (IQR), overall range (whiskers), and the presence of outliers, indicating how various states managed their inpatient bed capacities during the reporting period.

States with High Occupancy Rates: States like State9 have higher median occupancy rates, suggesting higher utilization of inpatient beds.

States with Lower Occupancy Rates: States like State1 have lower median occupancy rates, indicating relatively less strain on their inpatient bed capacity.


# Create boxplot for Weekly Average ICU Beds Occupied:

```{r}
ggplot(filtered_data, aes(x = `Geographic.aggregation`, y = `Weekly.Average.ICU.Beds.Occupied`)) +
  geom_boxplot(fill = "lightpink", color = "darkred") +
  coord_flip() +
  labs(title = "Boxplot of Weekly Average ICU Beds Occupied by State", x = "Geographic Aggregation", y = "Weekly Average ICU Beds Occupied")

```
Description of the Plot:

The boxplot visualizes the Weekly Average ICU Beds Occupied by state. Here are the key observations:

1.X-axis: The Weekly Average ICU Beds Occupied.

2.Y-axis: The Geographic Aggregation, representing different selected states.

This boxplot provides a clear visual comparison of the weekly average ICU beds occupied across different states. It highlights the differences in central tendency (median), spread (IQR), overall range (whiskers), and the presence of outliers, indicating how various states managed their ICU bed capacities during the reporting period.

States with High ICU Bed Occupancy Rates: States like State6 have higher median occupancy rates, suggesting higher utilization of ICU beds.

States with Lower ICU Bed Occupancy Rates: States like State10 have lower median occupancy rates, indicating relatively less strain on their ICU bed capacity.

## Explanation and Comparison:

Generating Synthetic Data:

The synthetic data is generated with the same structure and similar statistical properties as the original data. This includes date ranges, geographic aggregation categories, and numeric ranges for key metrics.

The faux package is used to create the synthetic data with the required number of observations and similar distribution characteristics.

## Similarities:

The EDA on synthetic data involves summarizing key variables and plotting distributions and boxplots to understand central tendencies and variability.

The synthetic data plot closely mirrors the distribution characteristics of the original data plot, particularly in terms of the right-skewed distribution and concentration of lower values for Weekly Average COVID-19 Admissions .

The histograms show the distribution of weekly averages for COVID-19 admissions, inpatient beds occupied, and ICU beds occupied. They exhibit similar patterns to the original data, with right-skewed distributions.

The boxplots provide insights into variability across different states, with central tendencies and outliers indicating similarities and differences in hospital metrics.




